Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-12 paper/paper.tex
split formula 32
compacting formulae
removed useless omega-argument in densities

Kullback-Leibler divergence of density $q$ from $p$ is
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
and is the natural choice for $\rho$, the main reason being that it satisfies the following
identity for well-behaved densities:
\begin{equation}  \label{eq:kl-div-master}
  \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q(\omega) \| \pi(\omega))
    + KL(q(\omega) \| p(\omega \mid D))
    \,.
\end{equation}
This identity yields an equivalent problem for \eqref{eq:variational-progam}:
maximizing the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
\begin{aligned}
  & \underset{\theta, \phi, \lambda}{\text{maximize}}
    & \mathcal{L}(\theta; \phi, \lambda)
      & = \mathbb{E}_{\omega \sim q_{\theta}}
          \log p_{\phi}(D \mid \omega)
        - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
        % \mathbb{E}_{\omega \sim q_{\theta}}
          % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % \\
  % & &
  %     & = \mathbb{E}_{\omega \sim q_{\theta}}
  %         \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %       + \mathbb{H}(q_{\theta})
  \,,
\end{aligned}
\end{equation}
where the prior and likelihood are allowed to depend on explicit parameters. The most
commonly used algorithm to optimize \eqref{eq:elbo_general} is EM-algorithm. Other
variational objectives are possible, provided $p(\omega \mid D)$ is probed only through
$\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2018}.

\textit{Stochastic Gradient Variational Bayes} (SGVB), proposed by \citep{kingma_auto-encoding_2014},
is a differentiable unbiased Monte-Carlo estimator of \eqref{eq:elbo_general}, which makes
it possible to employ stochastic gradient-based optimization methods. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and differentiable $
  g(\varepsilon; \theta)
$, then the SGVB is
\begin{equation}  \label{eq:sgvb}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid \omega_{il})
          \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta}(\omega) \| \pi(\omega))
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. Intractable KL-divergence terms can be replaced
by differentiable unbiased Monte-Carlo estimators, \citep{kingma_auto-encoding_2014}.
%
For the purposes of this study, we assume that the observed data is independent, and
use minibatch version of \eqref{eq:sgvb} with $L=1$:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta}(\omega) \| \pi_{\lambda}(\omega))
      + \frac{N}{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,,
\end{equation}
for a random subsample $(z_i)_{i=1}^M$ from $D$ and iid $
  (\varepsilon_i)_{i=1}^M \sim p_\varepsilon
$.
%
\todo{DSVI}

@@

With special family of posterior approximation, variational inference can be used as
a regularization method and as model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect, \citep{wan_regularization_2013},
and Gaussian dropout, \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and proposes \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. This unravels the divergence term in \eqref{eq:elbo} into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression for \verify{truncated} $\alpha \in (0, 1)$. \citet{molchanov_variational_2017}
refine the approximation by using sigmoid and soft-plus basis functions.
%
In sec.~\ref{sub:real-chisq-grad} we verify the approximation against Monte-Carlo
estimates of \eqref{eq:improper-kl-div-real} for $\alpha$ varying over a fine logarithmic
grid, and confirm its accuracy, by comparing its derivative to the exact derivative of
\eqref{eq:improper-kl-div-real} with respect to $\log \alpha$.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates

Flexibility in specification of $q_\theta$ enables structured dropout with shared
$\alpha$ across groups of parameters $\omega$. \citet{molchanov_variational_2017} study
Variational Dropout from a model sparsification perspective, optimizing $\alpha$ for each
individual parameter. Another variant of Bayesian dropout, called \textit{Automatic Relevance
Determination}, is developed in \citep{kharitonov_variational_2018}. The key difference is
that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes}
\citep{citation_needed}, fits a prior distribution while performing variational inference.
Maximizing \eqref{eq:elbo} over $\lambda$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta(\omega) \| \pi(\omega))
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

We consider two priors for Bayesian dropout: log-uniform and $\cplx$-Gaussian empirical
Bayes, \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q(\omega) \| \pi(\omega))
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = \sum_{ij} \bigl(
        - \mathbb{H}(q(\omega_{ij}))
        - \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
      \bigr)
    \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ below for brevity.

@@

\begin{equation}  \label{eq:g-m-deriv-int-2}

@@

@@

    = x^{-\tfrac{m}2} e^{-x}

>>>>

Kullback-Leibler divergence of density $q$ from $p$ is
\begin{equation}  \label{eq:kl-div-def}
  KL(q \| p)
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q}
      \log{q(\omega)}
    - \mathbb{E}_{\omega \sim q}
      \log{p(\omega)}
    \,,
\end{equation}
% and, despite it not being a metric and requiring distributions with nested support,
and is the natural choice for $\rho$, the main reason being that it satisfies the following
identity for well-behaved densities:
\begin{multline}  \label{eq:kl-div-master}
  % \log p(D)
    % = \mathbb{E}_{q(\omega)} \log \frac{p(D, \omega)}{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{p(\omega \mid D)}
    % = \mathbb{E}_{q(\omega)} \log \frac{
    %   p(D \mid \omega) \pi(\omega)
    % }{q(\omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
    % - \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{\pi(\omega)}
    % = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    % + \mathbb{E}_{q(\omega)} \log \frac{q(\omega)}{p(\omega \mid D)}
  \log p(D)
    - KL(q \| p(\omega \mid D))
    \\ = \mathbb{E}_{q(\omega)} \log{p(D \mid \omega)}
    - KL(q \| \pi)
    \,.
\end{multline}
This identity yields an equivalent problem for \eqref{eq:variational-progam}:
maximizing the \textit{evidence lower bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \mathbb{E}_{\omega \sim q_{\theta}}
        \log p_{\phi}(D \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where the prior and likelihood are allowed to depend on explicit parameters. The most
commonly used algorithm to optimize \eqref{eq:elbo_general} is EM-algorithm. Other
variational objectives are possible, provided $p(\omega \mid D)$ is probed only through
$\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2018}.

\textit{Stochastic Gradient Variational Bayes} (SGVB), proposed by \citep{kingma_auto-encoding_2014},
is a differentiable unbiased Monte-Carlo estimator of \eqref{eq:elbo_general}, which makes
it possible to employ stochastic gradient-based optimization methods. If $
  \omega \sim q_{\theta}
$ is equivalent in distribution to $
  \omega = g(\varepsilon; \theta)
$ for some non-parametric random variable $
    \varepsilon \sim p_\varepsilon
$ and differentiable $
  g(\varepsilon; \theta)
$, then the SGVB is
\begin{equation}  \label{eq:sgvb}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      % + \frac{N}{\lvert B \rvert} \sum_{i\in B}
        % \log p(z_i \mid \omega_i)b
      + \frac1{L} \sum_{l=1}^L
        \log p_{\phi}(D \mid g(\varepsilon_{l}; \theta))
        % \log p_{\phi}(D \mid \omega_{l})
        %   \Big\vert_{\omega_{l} = g(\varepsilon_{l}; \theta)}
      % =
      % - KL(q_{\theta} \| \pi)
      % + N \mathbb{E}_{B \sim \{D\}_M \otimes q_\theta^M}
        % \hat{\mathbb{E}}_{z,\omega \sim B}
          % \log p(z \mid \omega)
    \,, 
\end{equation}
where $
  (\varepsilon_{l})_{l=1}^L
$ is an iid sample from $p_\varepsilon$. Intractable KL-divergence terms can be replaced
by differentiable unbiased Monte-Carlo estimators, \citep{kingma_auto-encoding_2014}.
%
For the purposes of this study, we assume that the observed data is independent, and
use minibatch version of \eqref{eq:sgvb} with $L=1$:
\begin{equation}  \label{eq:elbo}
  \widetilde{\mathcal{L}}(\theta; \phi, \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \frac{N}{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,,
\end{equation}
for a random subsample $(z_i)_{i=1}^M$ from $D$ and iid $
  (\varepsilon_i)_{i=1}^M \sim p_\varepsilon
$.
%
\todo{DSVI}

@@

With special family of posterior approximation, variational inference can be used as
a regularization method and as model sparsification technique. \citet{kingma_variational_2015}
consider Dropout \citep{hinton_improving_2012}, DropConnect, \citep{wan_regularization_2013},
and Gaussian dropout, \citep{srivastava_dropout_2014,wang_fast_2013} through the lens
of Bayesian inference methods and proposes \textit{Variational Dropout}. They argue that
the multiplicative noise introduced by these methods induces a distribution equivalent to
a fully factorized variational posterior of the form $
  q_\theta(\omega) = \prod_j q_{\theta}(\omega_j)
$, where $q_{\theta}(\omega_j)$ is $\omega_j = \mu_j \xi_j$ with $
  \xi_j \sim p_\theta(\xi_j)
$ iid from some $p_\theta(\xi)$.
%
% Binary dropout with rate $p \in (0, 1)$ uses $
%   p_\xi
%     = \mathcal{B}\bigl(
%       \{0, \tfrac1{1-p}\}, 1-p
%     \bigr)
% $, whereas in Gaussian dropout $
%   \xi \sim \mathcal{N}(1, \alpha)
% $ for $\alpha = \tfrac{p}{1-p}$.
% This suggests making $\alpha$ a variational parameter
% and optimizing it in \eqref{eq:elbo_general} with an appropriate penalty term. Thus,
%
Variational Dropout assumes fully factorized Gaussian approximate posterior $
  q_\theta(\omega)
    = \prod_j \mathcal{N}(\mu_j, \alpha_j \mu_j^2)
$ and factorized log-uniform prior $\pi(\omega)$ with $
  \pi(\omega_j) \propto \lvert \omega_j \rvert^{-1}
$. This unravels the divergence term in \eqref{eq:elbo} into a sum of the following
analytically intractable terms:
\begin{equation}  \label{eq:improper-kl-div-real}
  KL(q_\theta \| \pi)
    \propto
      \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximate this expression by a non-linear polynomial
regression for \verify{truncated} $\alpha \in (0, 1)$. \citet{molchanov_variational_2017}
refine the approximation by using sigmoid and soft-plus basis functions.
%
In sec.~\ref{sub:real-chisq-grad} we verify the approximation against Monte-Carlo
estimates of \eqref{eq:improper-kl-div-real} for $\alpha$ varying over a fine logarithmic
grid, and confirm its accuracy, by comparing its derivative to the exact derivative of
\eqref{eq:improper-kl-div-real} with respect to $\log \alpha$.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates

Flexibility in specification of $q_\theta$ enables structured dropout with shared
$\alpha$ across groups of parameters $\omega$. \citet{molchanov_variational_2017} study
Variational Dropout from a model sparsification perspective, optimizing $\alpha$ for each
individual parameter. Another variant of Bayesian dropout, called \textit{Automatic Relevance
Determination}, is developed in \citep{kharitonov_variational_2018}. The key difference is
that the log-uniform factor $\pi(\omega_{ij})$ is replaced by a proper Gaussian prior $
  \mathcal{N}(0, \tau^{-1}_{ij})
$ with learnable precision $\tau_{ij} > 0$. This method, known as \textit{Empirical Bayes}
\citep{citation_needed}, fits a prior distribution while performing variational inference.
Maximizing \eqref{eq:elbo} over $\lambda$, holding other parameters fixed, yields $
  \tau^* = {(\mu^2 + \sigma^2)}^{-1}
$, whence
\begin{equation}  \label{eq:ard-kl-div-real}
  KL(q_\theta \| \pi)
    = \log{\bigl(1 + \tfrac1{\alpha} \bigr)}
    \,.
\end{equation}

@@

We consider two priors for Bayesian dropout: log-uniform and $\cplx$-Gaussian empirical
Bayes, \citep{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}.
For a fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = \sum_{ij} \bigl(
        - \mathbb{H}(q(\omega_{ij}))
        - \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
      \bigr)
    \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ below for brevity.

@@

\begin{align}  \label{eq:g-m-deriv-int-2}

@@

    &= x^{-\tfrac{m}2} e^{-x} \sum_{n\geq 0}

@@

    &= x^{-\tfrac{m}2} e^{-x}
